---
title: "Technical Exercise"
author: "Dennis Valentine"
date: "2024-01-11"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Libraries ---------------------------------------------------------------

library(dplyr)
library(ggplot2)
```

```{r loading_data}
md <- read.csv(file = "data/MedicalDictionary.csv")


patient <- read.csv(file = "data/Patient.csv") %>% 
  select(-X) %>% # cleaning data 
  filter(acceptable == 1) %>% # QC
  unique() %>% #remove duplicate rows
  mutate(regstartdate = as.Date(regstartdate), regenddate = as.Date(regenddate )) %>%  
  filter(regstartdate >= as.Date("1945-01-01") & regstartdate <= as.Date("2023-01-01")) %>%   # removing mistakes in the regdate 
  filter( (regenddate >= as.Date("1945-01-01") & regenddate <= as.Date("2023-01-01")) | is.na(regenddate)) %>%  # NA here could be active patients -- make sure no one las left before joiing
  filter( ! (regenddate <=  regstartdate) | is.na(regenddate)) %>% 
  group_by(patid) %>% 
  add_count()

patient %>% 
  arrange(patid) %>% 
  filter(n > 1) # YOB is different, gender is different -- this causes problems later on. At this point I have to leave the data in - one way this can be addressed is if it can be linked to xternal data sets like HES to establish the geneder and YOB.   



obvs <- read.csv(file = "data/Observation.csv") %>% 
  select(-X) %>% # medcodeid links to the dictionary 
  mutate(enterdate  = as.Date(enterdate)) %>% 
  filter(enterdate >= as.Date("1945-01-01") & enterdate <= as.Date("2023-01-01"))   # removing mistakes  
# sum(obvs$obsid == "") # checking the data 
# sum(is.na(obvs$obsid))


problem <- read.csv(file = "data/Problem.csv")
problem %>% head()

# Consultation.csv
cons <- read.csv(file = "data/Consultation.csv") %>% 
  unique()
cons %>% head() 

# Practice.csv
practice <- read.csv(file = "data/Practice.csv")
practice

```

## Joins
Join across the tables to generate a table or dataframe with the following information. In comments, explain how you dealt with any inconsistencies in the data.

```{r cars}

p1 <- patient %>% # I've already cleaned up the data as I was exporing it. 
  transmute(patid, age_at_2023 = 2023 - as.integer(yob), gender, regstartdate,  regenddate) %>% 
  unique() # try to remove multiple patid 


df <- left_join(x = p1, y = cons %>% select(patid:pracid), by = "patid") %>% # 1: many mapping, -- gender issues with FyHqc6 -- leave them in right now, no real way to know what the true gender is without additional data e.g. link to HES
  left_join(y =  obvs, by = c("patid", "consid", "pracid")) %>%  # obvs[30817,] == FyHqc6 person ID causing issues again. most likely more people in here are duplicates with issues but gobal QC on the data set is out of scope
  select(-obsdate ) %>% # this is subjective - people might not recall correctly, skewing answers. But then again if they don't get an appointment for weeks then the entry date is also skewed. I have to make a call because we don't have this in THIN. Furture work - plot histogram between two date types. Not sure what Aurum's normal practice is here  
  left_join(y = md, by = c("medcodeid" = "aurum_code")) %>% # dates causing issues -- just take the minimum data and again push for global QC 
   # not asked for term but this is one of those times I'll do more because from experience, clients don't always know what they want. They need something human-readable. Won't break protocol if locked in a contract or approved by some review committee 
  select(-pracid)

df %>% summary() # someone is aged 118 in 2023?? 

# In a database as big as CPRD anything that can go wrong would have, at some level, gone wrong. I think this data set was similar but I didn't explore it that well due to the time limit. 


```

## Counts

Number of consultations for each patient (save as .csv, and print number for patient 02A27z)

```{r counts1}
cons_per_patient <- df %>% 
  filter(!is.na(consid)) %>% 
  count(patid, consid) %>% 
  add_count(patid)  %>% 
  distinct(patid, n_consultations = nn) 

write.csv(x = cons_per_patient, file = "data/cons_per_patient.csv", row.names = FALSE)

cons_per_patient %>% 
  filter(patid == "02A27z")


```

Number of observations for each patient (save as .csv, and print number for patient 02A27z)
```{r, counts2}
obvs_per_patient <- df %>% 
  filter(!is.na(medcodeid )) %>% 
  count(patid, medcodeid  ) %>% 
  add_count(patid)  %>% 
  distinct(patid, n_obvs = nn)

write.csv(x = obvs_per_patient, file = "data/obvs_per_patient.csv", row.names = FALSE)

df %>% 
  filter(patid == "02A27z") %>% 
  filter(!is.na(medcodeid )) %>% 
  count(medcodeid) %>% 
  nrow()

```

Mean number of observations per consultation
```{r mean}
n_obvs_per_con <- df %>% 
  ungroup() %>% 
  filter(!is.na(medcodeid) & !is.na(consid)) %>% 
  distinct(consid, medcodeid) %>% 
  group_by(consid) %>% 
  add_count(name = "n_consid") %>% 
  distinct(consid, n_consid) %>% 
  ungroup() %>% 
  summarise(mean_obvs = mean(n_consid))
# mean = 3.46 observations (i.e. 4 observations per consultation)

```

Please also display number of consultations and observations per patient as a histogram.
```{r histogram}

pat_cons <- df %>% 
  distinct(patid, consid) %>% 
  filter(!is.na(consid)) %>% 
  summarise( n_cons = n())
  
pat_obvs <-   df %>% 
  distinct(patid, medcodeid) %>% 
  filter(!is.na(medcodeid)) %>% 
  summarise( n_obvs = n())
  
plotting_df <- full_join(x = pat_obvs, y = pat_cons, by = "patid")
summary(plotting_df)

ggplot(data = plotting_df) +
  geom_histogram(aes(x = n_obvs), bins = 105, fill = "blue") +
  geom_histogram(aes(x = n_cons), bins = 105, fill = "red" ) +
  ggtitle(label = "Histogram showing consultations and observations per patient", subtitle = "blue = number of consultations, red = number of observation") +
  xlab("")
  
# I should have done it differently - I should have done it on 2 different histograms and glued it together with `cowplot` or similar package - the scales are well off!! Not good :( 


## If I had more time to redo the plot it would be something like this: 
#devtools::install_github("johannesbjork/LaCroixColoR")
# LaCroixColoR::lacroix_palettes$paired # my fav colour palettes
library(cowplot)
pat_cons_plot <- ggplot(data = pat_cons, aes(x = n_cons)) +
  geom_histogram(bins = 31, fill = "#C70E7B") +
  xlim(10, 47) +
  theme_minimal() +
  ggtitle(label = "Histogram of consultations per person") +
  xlab(label = "Number of consultations") +
  ylab(label = "Number of people")

pat_obvs_plot <- ggplot(data = pat_obvs, aes(x = n_obvs)) +
  geom_histogram(bins = 68, fill = "#007BC3") +
  xlim(35, 110) +
  theme_minimal() +
  ggtitle(label = "Histogram of observations per person") +
  xlab(label = "Number of observations") +
  ylab(label = "Number of people")

plot_grid(pat_cons_plot, pat_obvs_plot, labels = c('A', 'B'), label_size = 12)


```
  
3. Identify how many patients have each of the following conditions:
    - Migraine
    - Type 2 diabetes
    - Stomach ulcer  
    
```{r phenotyping}


# Phenotyping is subjective so code lists are shared in databases like:
# UCL/HDR-UK: https://phenotypes.healthdatagateway.org/
# LSHTM: https://datacompass.lshtm.ac.uk/view/keywords/Code_list.html
# Cambridge: https://www.phpc.cam.ac.uk/pcu/research/research-groups/crmh/cprd_cam/codelists/v11/
# Birmingham
# QOF https://digital.nhs.uk/data-and-information/data-collections-and-data-sets/data-collections/quality-and-outcomes-framework-qof/quality-and-outcome-framework-qof-business-rules/quality-and-outcomes-framework-qof-business-rules-v47.0-2022-2023

# If this was a normal coding system like Read or ICD then I'd phenotype based
# on existing publications and cite the paper. I would also get green light
# from the client for projects


migrane_c <- md %>% 
  filter(grepl(pattern = "Migraine", x = Term, ignore.case = TRUE)) %>% 
  select(aurum_code) %>% 
  unlist(use.names = FALSE)

t2d_c <- md %>%  # known different subtypes. 
  filter(grepl(pattern = "diabetes", x = Term, ignore.case = TRUE)) %>% 
  filter(grepl(pattern = "1", x = Term, ignore.case = TRUE)) %>% 
  select(aurum_code) %>% 
  unlist(use.names = FALSE)


su_c <- md %>% 
  filter(grepl(pattern = "Stomach|ulcer", x = Term, ignore.case = TRUE)) %>% 
  filter(aurum_code == "NDWL524827") %>%  # on;y 1 code 
  select(aurum_code) %>% 
  unlist(use.names = FALSE)

# since i'm running the same code >=3 times I'll write a function
count_disease <- function(x){
  x %>% 
    add_count() %>% 
    filter(n > 1) %>% 
    distinct(patid) %>% 
    nrow() %>% 
    print()
}



df %>% 
  filter(medcodeid %in% migrane_c) %>% 
  count_disease() # 20 distinct patients 

df %>% 
  filter(medcodeid %in% t2d_c) %>% 
  count_disease() # 15

df %>% 
  filter(medcodeid %in% su_c) %>% 
  count_disease() # 27

```

## Stats
Choose one of the above conditions. Choose an appropriate approach to statistically test if there are gender differences in the presence of this condition (1 = Male, 2 = Female, 3 = Unspecified/Other) and show your output.

```{r stats}
cases <- df %>% 
  filter(medcodeid %in% su_c) %>% 
  add_count() %>% 
  filter(n > 1) %>% 
  distinct(patid) %>% 
  mutate(status = 1)

test_df <- left_join(x = df %>% distinct(patid, gender), y = cases, by = "patid") %>% 
  mutate(status = ifelse(test = is.na(status), yes = 0, no = status))

sum(test_df$status) # these must be issues with the gender again 
test_table <- table(gender = test_df$gender, disease_status = test_df$status)
chisq.test(test_table)
```

## Version control & dependency control
We do not expect you to provide robust version control procedures or dependency control for this brief exercise. However, please explain how you would do so given the appropriate time and resoure.

I use git for version control - GitHub for personal use and my PhD while BitBucket for my professional life at Cegedim. Dependencey control could be done in a few different ways. I experimented with packrat before I discovered Docker images. I've build a few containers over at dockerhub. Here is a link:
https://hub.docker.com/repository/docker/dendendocks/c3-olap-1536/general. In this instance, as it was part of my PhD, the docker file isn't publicly
available.
  